---
title: "Advanced R Study Notes"
author: "Chen Qinqqing"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document:
    fig_width: 12
    fig_height: 10
    fig_caption: true
    toc: true # table of content true
    toc_depth: 4  # upto three depths of headings (specified by #, ## and ###)
    number_sections: FALSE  ## if you want number sections at each table header
    theme: united  
    highlight: tango  # specifies the syntax highlighting style
    toc_float: true   # you can add your custom css, should be in same folder
    code_folding: show
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Foundations
## Names and Values


Use the [lobstr](https://github.com/r-lib/lobstr) package to dig into the internal representation of R objects.

Useful Links: 

  - [memory profiling](https://cran.r-project.org/doc/manuals/R-exts.html#Profiling-R-code-for-memory-use)
  - [SEXPs](https://cran.r-project.org/doc/manuals/R-ints.html#SEXPs)
  
```{r warning=FALSE}
#install.packages("lobstr")
library(lobstr)
```

### Binding basics

```{r}
x <- c(1, 2, 3)
y <- x
##identifier
obj_addr(x)
obj_addr(y)
```


### Copy-on-modify

```{r}
x <- c(1, 2, 3)
y <- x
cat(tracemem(x), "\n")
cat(tracemem(y), "\n")
y[[3]] <- 5L
untracemem(y)
cat(tracemem(y), "\n")
```

```{r}
f <- function(a) {
  a
}

x <- c(1, 2, 3)
cat(tracemem(x), "\n")
#> <0x55f3108>

z <- f(x)
# there's no copy here!

untracemem(x)
```

```{r}
l1 <- list(1, 2, 3)
l2 <- l1
l2[[3]] <- 4
ref(l1, l2)
```

```{r}
d1 <- data.frame(x = c(1, 5, 6), y = c(2, 4, 3))
d2 <- d1
d2[, 2] <- d2[, 2] * 2
d3 <- d1
d3[1, ] <- d3[1, ] * 3
tracemem(d3)
```

```{r}
x <- c("a", "a", "abc", "d")
ref(x, character = TRUE)
```

### Object size 

```{r}
obj_size(letters)
#> 1,712 B
obj_size(ggplot2::diamonds)
#> 3,456,344 B

x <- runif(1e6)
obj_size(x)
#> 8,000,048 B

y <- list(x, x, x)
obj_size(y)
#> 8,000,128 B
obj_size(list(NULL, NULL, NULL))

banana <- "bananas bananas bananas"
obj_size(banana)
#> 136 B
obj_size(rep(banana, 100))
#> 928 B


obj_size(x, y)

obj_size(1:3)
#> 680 B
obj_size(1:1e3)
#> 680 B
obj_size(1:1e6)
#> 680 B
obj_size(1:1e9)
#> 680 B1e3
```


### Modify-in-place

#### Objects with a single binding
Modifying an R object usually creates a copy. There are two exceptions:

  - Objects with a single binding get a special performance optimisation.
  - Environments, a special type of object, are always modified in place.

```{r}
x <- data.frame(matrix(runif(5 * 1e4), ncol = 5))
medians <- vapply(x, median, numeric(1))

for (i in seq_along(medians)) {
  x[[i]] <- x[[i]] - medians[[i]]
}

cat(tracemem(x), "\n")

for (i in 1:5) {
  x[[i]] <- x[[i]] - medians[[i]]
}
```


#### Environments 

```{r}
e1 <- rlang::env(a = 1, b = 2, c = 3)
e2 <- e1
e2$c
e1$c <- 4
e2$c

e <- rlang::env()
e$self <- e

ref(e)
```

### Unbinding and the garbage collector

```{r}
x <- 1:3
x <- 2:4
rm(x)
```


If you want to find out when the GC runs, call gcinfo(TRUE) and GC will print a message to the console every time it runs.

You can force garbage collection by calling `gc()`. But despite what you might have read elsewhere, there’s never any need to call `gc()` yourself. The only reasons you might want to call `gc()` is to ask R to return memory to your operating system so other programs can use it, or for the side-effect that tells you how much memory is currently being used:

```{r}
gc() 
mem_used()
```















