---
title: "Efficient R Programming"
author: "Chen Qinqqing"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document:
    toc: true 
    toc_depth: 3  
    number_sections: FALSE  
    theme: united  
    highlight: tango  
    toc_float: true   
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Chapter 1: Introduction 
## Benchmarking and profiling
### Benchmarking
Modifying elements from one benchmark to the next and recording the results after the alteration enables us to determine the fastest piece of code. Benchmarking is important in the efficient programmer’s tool-kit: you may think that your code is faster than mine but benchmarking allows you to prove it. The easiest way to benchmark a function is to use `system.time()`. However it is important to remember that we are taking a sample. We wouldn’t expect a single person in London to be representative of the entire UK population, similarly, a single benchmark provides us with a single observation on our functions behaviour. Therefore, we’ll need to repeat the timing many times with a loop.

An alternative way of benchmarking, is via the flexible **microbenchmark** package. This allows us to easily run each function multiple times (by default  100), enabling the user to detect microsecond differences in code performance. We then get a convenient summary of the results: the minimum/maximum, lower/upper quartiles and the mean/median times. We suggest focusing on the median time to get a feel for the standard time and the quartiles to understand the variability

Standard units are:
  - milliseconds (ms), one thousand functions takes a second;
  - microseconds ( μs), one million function calls takes a second;
  - nanoseconds (ns), one billion calls takes a second

```{r benchmark_1}
library(microbenchmark)
df = data.frame(v = 1:4, name = letters[1:4])
microbenchmark(df[3,2], df[3, "name"], df$name[3])
```

```{r benchmark_2}
x = 1:100 # initiate vector to cumulatively sum
# Method 1: with a for loop (10 lines)
cs_for = function(x){
  for(i in x){
    if(i == 1){
      xc = x[i]
    } else {
      xc = c(xc, sum(x[1:i]))
      }
  }
  xc
}

# Method 2: with apply (3 lines)
cs_apply = function(x){
  sapply(x, function(x) sum(1:x))
}

# Method 3: cumsum (1 line, not shown)
microbenchmark(cs_for(x), cs_apply(x), cumsum(x))

# Test how long it takes to subset the data frame 50,000 times:
system.time(
  for(i in 1:50000){
    df[3, 2]
  }
)
```

### Profiling
Benchmarking generally tests the execution time of one function against another. Profiling, on the other hand, is about testing large chunks of code.

```{r profvis}
library("profvis")
# profvis(expr = {
#   # Stage 1: load packages
#   # library("rnoaa") # not necessary as data pre-saved
#   library("ggplot2")
#   
#   # Stage 2: load and process data
#   out = readRDS("extdata/out-ice.Rds")
#   df = dplyr::rbind_all(out, id = "Year")
#   
#   # Stage 3: visualise output
#   ggplot(df, aes(long, lat, group = paste(group, Year))) +
#     geom_path(aes(colour = Year)) 
#   ggsave("figures/icesheet-test.png")
# }, interval = 0.01, prof_output = "ice-prof")
```


# Chapter 2: Efficient set-up
This chapter describes the set-up that will enable a productive workflow. It explores how the operating system, R version, startup files and IDE can make your R work faster. Understanding and at times changing these set-up options can have many knock-on benefits.  It covers the following topics:
  - R and the operating systems: system monitoring on Linux, Mac and Windows
  - R version: how to keep your base R installation and packages up-to-date
  - R start-up: how and why to adjust your .Rprofile and .Renviron files
  - RStudio: an integrated development environment (IDE) to boost your programming productivity
  - BLAS and alternative R interpreters: looks at ways to make R faster
  
## Top 5 tips for an efficient R set-up
- Use system monitoring to identify bottlenecks in your hardware/code.
- Keep your R installation and packages up-to-date.
- Make use of RStudio’s powerful autocompletion capabilities and shortcuts.
- Store API keys in the .Renviron file.
- Use BLAS if your R number crunching is too slow.

```{r load_library_1}
library(benchmarkme)
```

## Operating system
```{r system_info}
Sys.info()
```


### Operating system and resource monitoring
Resource monitoring is the process of checking the status of key OS variables. For computationally intensive work, it is sensible to monitor system resources in this way. Resource monitoring can help identify computational bottlenecks. Alongside R profiling functions such as **profvis**, system monitoring provides a useful tool for understanding how R is performing in relation to variables reporting the OS state, such as how much RAM is in use, which relates to the wider question of whether more is needed.

CPU resource allocated over time is another common OS variable that is worth monitoring. A basic use case is to check whether your code is running in parallel, and whether there is spare CPU capacity on the OS that could be harnessed by parallel code.

System monitoring is a complex topic that spills over into system administration and server management. Fortunately there are many tools designed to ease monitoring on all major operating systems.
  - On Linux, the shell command top displays key resource use figures for most distributions. htop and Gnome’s System Monitor (gnome-system-monitor, see figure 2.1) are more refined alternatives which use command-line and graphical user interfaces respectively. A number of options such as nethogs monitor internet usage.
  - On Mac the Activity Monitor provides similar functionality. This can be initiated from the Utilities folder in Launchpad.
  - On Windows the Task Manager provides key information on RAM and CPU use by process. This can be started in modern Windows versions by typing Ctrl-Alt-Del or by clicking the task bar and ‘Start Task Manager’.

## R version
### Installing R
- On Windows, a single .exe file (hosted at cran.r-project.org/bin/windows/base/) will install the base R package.
- On a Mac, the latest version should be installed by downloading the `.pkg` files hosted at [cran.r-project.org/bin/macosx/]("https://cran.r-project.org/bin/macosx/").
- On Linux, the installation method depends on the distribution of Linux installed, althogh the principles are the same. We’ll cover how to install R on Debian-based systems, with links at the end for details on other Linux distributions. First stage is to add the CRAN repository, to ensure that the latest version is installed. If you are running Ubuntu 16.04, for example, append the following line to the file /etc/apt/sources.list: ```deb http://cran.rstudio.com/bin/linux/ubuntu xenial/```

Once the appropriate repository has been added and the system updated (e.g. with sudo apt-get update, r-base and other r- packages can be installed using the apt system. The following two commands, for example, would install the base R package (a ‘bare-bones’ install) and the package rcurl, which has an external dependency:
- `sudo apt-get install r-cran-base # install base R`
- `sudo apt-get install r-cran-rcurl # install the rcurl package`

### Updating R
- On Ubuntu-based systems, new versions of R should be automatically detected through the software management system, and can be installed with apt-get upgrade.
- On Mac, the latest version should be installed by the user from the `.pkg` files mentioned above.
- On Windows installr package makes updating easy:`installr::updateR() `

### Installing R packages
```{r install_package}
pkgs = c("raster", "leaflet", "rgeos") # package names
install.packages(pkgs)
```

### Installing R packages with dependencies
- For Debian-based distributions such as Ubuntu
```apt-cache search r-cran- # search for available cran Debian packages```
```sudo apt-get-install r-cran-rgdal # install the rgdal package (with dependencies)```
- For Windows: eg.For example the Rtools package for compiling C/C++ code on Windows can be installed with the following command:
```installr::install.rtools()```

### Updating R packages
```update.packages() # update installed CRAN packages```

### An overview of R’s startup files
- `.Renviron`, the primary purpose of which is to set environment variables. These tell R where to find external programs and can hold user-specific information than needs to be kept secret, typically API keys.

- `.Rprofile` is a plain text file (which is always called .Rprofile, hence its name) that simply runs lines of R code every time R starts. 

When R starts (unless it was launched with --no-environ) it first searches for `.Renviron` and then `.Rprofile`, in that order. Although `.Renviron` is searched for first, we will look at `.Rprofile` first as it is simpler and for many set-up tasks more frequently useful. Both files can exist in three directories on your computer.

### The location of startup files
Files in three folders are important in this process:
- `R_HOME`, the directory in which R is installed. The etc sub-directory can contain start-up files read early on in the start-up process. Find out where your `R_HOME` is with the `R.home()` command.
- `HOME`, the user’s home directory. Typically this is `/home/username` on Unix machines or `C:\Users\username` on Windows (since Windows 7). Ask R where your home directory is with, `Sys.getenv("HOME")`.
- R’s current working directory. This is reported by `getwd()`.
- Edit file 
  - ```file.edit("~/.Rprofile") # edit .Rprofile in HOME```
  - ```file.edit(".Rprofile") # edit project specific .Rprofile```
- Editing the `.Renviron` file in the same locations will have the same effect. The following code will create a user specific `.Renviron` file (where API keys and other cross-project environment variables can be stored), without overwriting any existing file.
  - ```user_renviron = path.expand(file.path("~", ".Renviron"))```
  - ```file.edit(user_renviron) # open with another text editor if this fails```

# Chapter 3: Efficient programming 

```{r load_library2}
library("compiler")
library("memoise")
```

## Top 5 tips for efficient programming
- Be careful never to grow vectors.
- Vectorise code whenever possible.
- Use factors when appropriate.
- Avoid unnecessary computation by caching variables.
- Byte compile packages for an easy performance boost.

## General advice
A **golden rule** in R programming is to access the underlying C/Fortran routines as quickly as possible; the fewer functions calls required to achieve this, the better.

### Memory allocation
Another general technique is to be careful with memory allocation. If possible pre-allocate your vector then fill in the values. Never grow an object.

```{r test1}
method1 = function(n) {
  vec = NULL # Or vec = c()
  for(i in seq_len(n))
    vec = c(vec, i)
  vec
} #Method 1 creates an empty vector and gradually increases (or grows) the length of the vector

method2 = function(n) {
  vec = numeric(n)
  for(i in seq_len(n))
    vec[i] = i
  vec
} #Method 2 creates an object of the final length and then changes the values in the object by subscripting

method3 = function(n) seq_len(n) #Method 3 directly creates the final object
microbenchmark(times = 100, unit = "s", 
               method1(10), method2(10), method3(10))
```

## The apply family
- `apply`: Apply functions over array margins
- `by`:	Apply a function to a data frame split by factors
- `eapply`	Apply a function over values in an environment
- `lapply`	Apply a function over a list or vector
- `mapply`	Apply a function to multiple list or vector arguments
- `rapply`	Recursively apply a function to a list
- `tapply`	Apply a function over a ragged array

```{r apply_eg}
data(movies, package = "ggplot2movies")
ratings = movies[, 7:16]
ratings[1, ]
popular = apply(ratings, 1, nnet::which.is.max)
plot(table(popular))
```


# Chapter 4: Efficient workflow 
## Top 5 tips for efficient workflow
- Start without writing code but with a clear mind and perhaps a pen and paper. This will ensure you keep your objectives at the forefront of your mind, without getting lost in the technology.
- Make a plan. The size and nature will depend on the project but time-lines, resources and ‘chunking’ the work will make you more effective when you start.
- Select the packages you will use for implementing the plan early. Minutes spent researching and selecting from the available options could save hours in the future.
- Document your work at every stage: work can only be effective if it’s communicated clearly and code can only be efficiently understood if it’s commented.
- Make your entire workflow as reproducible as possible. knitr can help with this in the phase of documentation.

## A project planning typology
Appropriate project management structures and workflow depend on the type of project you are undertaking. The typology below demonstrates the links between project type and project management requirements.10
  - **Data analysis**. Here you are trying to explore datasets to discover something interesting/answer some questions. The emphasis is on speed of manipulating your data to generate interesting results. Formality is less important in this type of project. Sometimes this analysis project may only be part of a larger project (the data may have to be created in a lab, for example). How the data analysts interact with the rest of the team may be as important for the project’s success as how they interact with each other.
  - **Package creation**. Here you want to create code that can be reused across projects, possibly by people whose use case you don’t know (if you make it publicly available). The emphasis in this case will be on clarity of user interface and documentation, meaning style and code review are important. Robustness and testing are important in this type of project too.
  - **Reporting and publishing**. Here you are writing a report or journal paper or book. The level of formality varies depending upon the audience, but you have additional worries like how much code it takes to arrive at the conclusions, and how much output does the code create.
  - **Software applications**. This could range from a simple Shiny app to R being embedded in the server of a much larger piece of software. Either way, since there is limited opportunity for human interaction, the emphasis is on robust code and gracefully dealing with failure.

Based on these observations we recommend thinking about which type of workflow, file structure and project management system suits your projects best. Sometimes it’s best not to be prescriptive so we recommend trying different working practices to discover which works best, time permitting.11

There are, however, concrete steps that can be taken to improve workflow in most projects that involve R programming. Learning them will, in the long-run, improve productivity and reproducibility. With these motivations in mind, the purpose of this chapter is simple: to highlight some key ingredients of an efficient R workflow. It builds on the concept of an R/RStudio project, introduced in Chapter 2, and is ordered chronologically throughout the stages involved in a typical project’s lifespan, from its inception to publication:

  - **Project planning**. This should happen before any code has been written, to avoid time wasted using a mistaken analysis strategy. Project management is the art of making project plans happen.
  - **Package selection**. After planning your project you should identify which packages are most suitable to get the work done quickly and effectively. With rapid increases in the number and performance of packages it is more important than ever to consider the range of options at the outset. For example `*_join()` from dplyr is often more appropriate than `merge()`, as we’ll see in 6.
  - **Publication**. This final stage is relevant if you want your R code to be useful for others in the long term. 

```{r visualising_plans}
library("DiagrammeR")
# Define the Gantt chart and plot the result (not shown)
mermaid("gantt
        Section Initiation
        Planning           :a1, 2016-01-01, 10d
        Data processing    :after a1  , 30d")
```

- [How to trust a R package]("https://simplystatistics.org/2015/11/06/how-i-decide-when-to-trust-an-r-package/")











