---
title: "Geocomputation with R"
author: "Chen Qinqqing"
date: "13/09/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eg1}
library(leaflet)
popup = c("Robin", "Jakub", "Jannes")
leaflet() %>%
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>%
  addMarkers(lng = c(-3, 23, 11),
             lat = c(52, 53, 49), 
             popup = popup)
```

```{r load_library}
library(sf)
library(raster)
library(spData)
library(spDataLarge)
```

# Geographic data in R
## Vector data 
- The geographic vector model is based on points located within a coordinate reference system (**CRS**)

### An introduction to simple features 
- `sf` can represent all common vector geometry types (raster data classes are not supported by sf): points, lines, polygons and their respective ‘multi’ versions (which group together features of the same type into a single feature)

- `sf` incorporates the functionality of the three main packages of the `sp paradigm`, `sp` (E. Pebesma and Bivand 2018) for the class system, `rgdal` (Bivand, Keitt, and Rowlingson 2018) for reading and writing data, `rgeos` (Bivand and Rundel 2018) for spatial operations undertaken by GEOS, in a single, cohesive whole. 

```{r eg2}
names(world)
plot(world)
summary(world["lifeExp"])
world[1:2, 1:3]
```

### Why simple features
Simple features is a widely supported data model that underlies data structures in many GIS applications including QGIS and PostGIS. A major advantage of this is that using the data model ensures your work is cross-transferable to other set-ups, for example importing from and exporting to spatial databases.
  - Fast reading and writing of data
  - Enhanced plotting performance
  - sf objects can be treated as data frames in most operations
  - sf functions can be combined using %>% operator and works well with the tidyverse collection of R packages
  - sf function names are relatively consistent and intuitive (all begin with st_)

```{r eg3}
library(sp)
world_sp = as(world, Class = "Spatial")
world_sp
```


### Basic map making
Basic maps are created in sf with `plot()`
```{r eg4}
plot(world[3:6])
plot(world["pop"])
```

Plots are added as layers to existing images by setting `add = TRUE`
Note that the first plot must only have one facet for `add = TRUE` to work. If the first plot has a key, `reset = FALSE` must be used (result not shown):
```{r eg5}
world_asia = world[world$continent == "Asia", ]
asia = st_union(world_asia)
plot(world["pop"], reset = FALSE)
plot(asia, add = TRUE, col = "red")
```

### Base plot arguments
There are various ways to modify maps with sf’s `plot()` method. Because sf extends base R plotting methods `plot()`’s arguments such as m`ain =` — which specifies the title of the map — work with sf objects (see `?graphics::plot` and `?par`)

```{r eg6}
plot(world["continent"], reset = FALSE)
cex = sqrt
```





